Transfer learning between Sentinel-1 acquisition modes enhances \\ the few-shot segmentation of natural oil slicks in the Arctic

### 2025-04-04 latex/manuscript.tex
submission formatting
remove line numbering, etc

\author{
    Julien Vadnais\orcidlink{0009-0003-2680-8841},~\IEEEmembership{Graduate Student Member,~IEEE}, 
    Benjamin Aubrey Robson\orcidlink{0000-0002-4987-7378}, 
    Christian Haug Eide\orcidlink{0000-0003-4949-9917}, \\
    Rune Mattingsdal\orcidlink{0000-0003-0168-2405}, 
    Malin Johansson\orcidlink{0000-0003-0129-2239},~\IEEEmembership{Member,~IEEE} 
        % <-this % stops a space
\thanks{Manuscript received March 31, 2025. This research was carried out in the context of the InFluSe project (Research Council of Norway Project 353044).}

@@

% The paper headers
\markboth{\LaTeX\ PDF, IEEE GEOSCIENCE AND REMOTE SENSING LETTERS,~Vol.~14, No.~8, March~2025}%
{Shell \MakeLowercase{\textit{et al.}}: A Sample Article Using IEEEtran.cls for IEEE Journals}

@@

\maketitle
\linenumbers
\begin{abstract} 
    Natural seepage is a significant contributor to marine hydrocarbon inputs. 
    Remote and intermittent seeps are difficult to monitor in the field, yet oil slicks can be observed by spaceborne synthetic aperture radar (SAR) because they reduce backscatter, 
    % because they reduce backscatter,
    creating potential for automatic mapping. In mapping tasks like segmentation, deep learning models excel, albeit needing large amounts of labeled images.
    To deal with scarcity of labeled images, transfer learning is an approach commonly used in computer vision, though still underutilized in remote sensing. In the case of oil slicks, 
    differences between Sentinel-1 acquisition modes, such as the interferometric wide (IW) in the North Sea and extra wide (EW) in the Arctic, complicate direct model transfer.
    Here, we present a use-case where transfer learning enhances the segmentation of natural oil slicks. We used labeled slicks in IW images in the North Sea to pretrain a series of DeepLabv3 and SAM models. 
    These models were then fine-tuned on EW-labeled slicks from two documented Arctic seeps on which we have only limited observations. Our results show clear evidence that transfer learning improves few-shot 
    segmentation, notably in challenging images. Overall, few studies have addressed transfer learning between SAR acquisition modes. This work contributes to improved monitoring of poorly 
    understood or yet undiscovered hydrocarbon seeps. 
\end{abstract}

\nolinenumbers
\begin{IEEEkeywords}
    Natural seepage, oil slicks, SAR, deep learning, transfer learning, segmentation, DeepLab, SAM
\end{IEEEkeywords}

% Line numbering for review only
\linenumbers

@@

% both deep learning models benefited from transferring pretrained model weights, 
% with DeepLabv3 performing best overall, though SAM achieved top accuracy with 
% fewer training images. The analysis was validated using two documented 
% seeps in the Arctic, where our method could potentially identify additional, 
% undiscovered seepage sites. Overall, our findings highlight the potential of 
% transfer learning. This approach remains underutilized in remote sensing, 
% yet it is promising for addressing data scarcity, with practical applications 
% such as monitoring natural seepage in remote marine environments. 
\section*{Acknowledgments}
Julien Vadnais received financial support from the Doctoral Research Scholarship of the Fonds de recherche du Québec. 
The authors would like to thank the EU Copernicus programme for allowing open access to the Sentinel-1 archives.
This research was carried out as part of the InFluSe project (Research Council of Norway Project 353044).
\bibliographystyle{IEEEtran}
\bibliography{references.bib}

>>>>

\author{
    Julien Vadnais\orcidlink{0009-0003-2680-8841},~\IEEEmembership{Graduate Student Member,~IEEE}, 
    Benjamin Aubrey Robson\orcidlink{0000-0002-4987-7378}, 
    Christian Haug Eide\orcidlink{0000-0003-4949-9917}, \\
    Rune Mattingsdal\orcidlink{0000-0003-0168-2405}, 
    Malin Johansson\orcidlink{0000-0003-0129-2239},~\IEEEmembership{Member,~IEEE} 
        % <-this % stops a space
\thanks{Manuscript received March 4, 2025. This research was carried out in the context of the InFluSe project (Research Council of Norway Project 353044).}

@@

% The paper headers
\markboth{\LaTeX\ PDF, IEEE GEOSCIENCE AND REMOTE SENSING LETTERS,~Vol.~XX, No.~Y, <Month>~2025}%
{Shell \MakeLowercase{\textit{et al.}}: A Sample Article Using IEEEtran.cls for IEEE Journals}

@@

\maketitle

% \linenumbers %% Uncomment to activate line numbering
\begin{abstract} 
    Natural seepage is a significant contributor to marine hydrocarbon inputs. 
    Remote and intermittent seeps are difficult to monitor in the field, yet oil slicks can be observed by spaceborne synthetic aperture radar (SAR) because they reduce backscatter, 
    % because they reduce backscatter,
    creating potential for automatic mapping. In mapping tasks like segmentation, deep learning models excel, albeit needing large amounts of labeled images.
    To deal with scarcity of labeled images, transfer learning is an approach commonly used in computer vision, though still underutilized in remote sensing. In the case of oil slicks, 
    differences between Sentinel-1 acquisition modes, such as the interferometric wide (IW) in the North Sea and extra wide (EW) in the Arctic, complicate direct model transfer.
    Here, we present a use-case where transfer learning enhances the segmentation of natural oil slicks. We used labeled slicks in IW images in the North Sea to pretrain a series of DeepLabv3 and SAM models. 
    These models were then fine-tuned on EW-labeled slicks from two documented Arctic seeps on which we have only limited observations. Our results show clear evidence that transfer learning improves few-shot 
    segmentation, notably in challenging images. Overall, few studies have addressed transfer learning between SAR acquisition modes. This work contributes to improved monitoring of poorly 
    understood or yet undiscovered hydrocarbon seeps. 
\end{abstract}

\begin{IEEEkeywords}
    Natural seepage, oil slicks, SAR, deep learning, transfer learning, segmentation, DeepLab, SAM
\end{IEEEkeywords}

@@

\section*{Acknowledgments}
Julien Vadnais received financial support from the Doctoral Research Scholarship of the Fonds de recherche du Québec. 
The authors would like to thank the EU Copernicus programme for allowing open access to the Sentinel-1 archives.
This research was carried out as part of the InFluSe project (Research Council of Norway Project 353044).
\bibliographystyle{IEEEtran}
\bibliography{references.bib}
